Syntax-Based Concept Alignment for Domain-Specific Machine Translation

### 2021-05-11 paper/paper.tex
done with bibliography

\section{Introduction}
Grammar-based translation pipelines such as those based on Grammatical Framework (GF) have been successfully employed in domain-specific Machine Translation (MT) \cite{scaling_up_gf}.  
What makes these systems well suited to the task is the fact that, when we constrain ourselves to a specific domain, where precision is most often more important than coverage, they can provide strong guarantees in terms of grammatical correctness.

@@

This work is a new attempt in the same direction, enabled by two multilinguality-oriented grammar formalisms developed over the course of the last 25 years: Universal Dependencies (UD) \cite{ws-2019-universal} and Grammatical Framework (GF) \cite{scaling_up_gf}.

@@

UD parsing alone would then be sufficient to extract (or propagate) tree-to-tree alignments, but not to automate the generation of a morphologically-aware translation lexicon for a generative grammar. 
This is where GF comes into play: after correspondences are inferred from a parallel text, our proposed system is able to convert them to GF grammar rules, easy to embed in a domain-specific grammar but also making it possible to immediately evaluate the system by carrying out small-scale translation experiments using pre-existing grammatical constructions implemented in GF's Resource Grammar Library (RGL) \cite{dummy}. \todo{say what the RGL is} 
This is made possible by \texttt{gf-ud}, a conversion tool described in \cite{dummy} and \cite{dummy}.

@@

The pseudocode of the basic extraction algorithm is shown in Figure \ref{TODO}\todo{add pseudocode}.
The input is a list of priority-sorted \textit{alignment criteria}, i.e.rules to determine whether two dependency trees should be aligned with each other, which will be discussed in greater detail in Section \ref{criteria}, and UD tree representation of the pair of sentences to align.
From an implementation point of view, UD trees are, as in \texttt{gf-ud}, rose trees where each node represents a token, obtained from the CoNLL-U files produced by the UD parser.

@@

\paragraph{Part-Of-Speech equivalence}
CoNNL-U files provide information not only on the syntactic role of the tokens a sentence is composed of, but also on their grammatical categories, represented as Universal Part-Of-Speech (POS) tags \cite{dummy}. 
Intuitively, if the nodes of two trees in matching contexts have the same POS tags, the two trees are more likely to correspond to each other than if not.

This is especially true if we focus, for instance, solely on the open class words (in the current implementation, defined as in the UD documentation \cite{dummy}), thus ignoring prepositions, determiners, auxiliary verbs and all other function words, which tend to behave differently across different languages.

@@

\paragraph{Known translation divergence}
Parallel texts often present significant, systematic cross-linguistic grammatical distinctions. 
When this is the case, it is often straightforward to define alignment criteria based on recognizing the corresponding patterns.
If many distinctions of this kind are specific to particular language pairs or even stylistical, some of them occur independently of what languages are involved and have nothing to do with idiomatic usage or aspectual, discourse, domain or word knowledge. Drawing inspiration from \cite{dummy} and \cite{dummy}, we refer to them as \textit{translation divergences} and handle the most common ones via alignment criteria that are used by default. \todo{add a few examples (<= one per type of divergence)}

@@

\section{Evaluation} \label{evaluation}
% TODO: move around things to have something here (or outline structure of the section)
\subsection{Data}
Because we want part of our evaluation to be independent from the quality of UD parsing, some of the experiments are carried out using manually annotated data. 
To this purpose, we use a 100-sentences subset of the Parallel UD (PUD) corpus, a set of 1000 handcrafted\footnote{Annotation is sometimes semi-automatic, but all treebanks are at least manually checked.}\todo{not even, I fear. Check this} multilingual treebanks in CoNLL-U format created for the CoNLL 2017 shared task on Multilingual Parsing from Raw Text to Universal Dependencies \cite{dummy}.
PUD treebanks are available in over 20 languages, of which we selected English, Italian and Swedish. 
Using this data limits the amount of alignment errors that are due annotation issues to a minimum, even though a small number of inconsistencies and imprecisions is present even in this corpus.

When it comes to testing the program on raw text, we use two bilingual sentence-aligned corpora consisting of course plans from the Department of Mathematics and Computer Science of the University of Perugia (for English-Italian) and from the Department of Computer Science and Engineering (CSE) shared between the University of Gothenburg and the Chalmers University of Technology (for English-Swedish). 
For brevity, we will refer to these two datasets as to the DMI and the CSE corpora, or collectively as to the course plans datasets.
This data, available in the project repository, was collected and sentence-aligned specifically for this work and a related Bachelor's thesis project \cite{dummy}.
When using raw text, our parser of choice is UDPipe \cite{dummy}. 
In particular, we use the ParTUT English and Italian models for the DMI corpus and models trained on the bilingual LinES English-Swedish treebank for the CSE corpus \cite{dummy}.

@@

We assess the correctness of the alignments the CE module is able to extract first from manually annotated data, then from the DMI and CSE corpus, comparing our results with those obtained with a standard statistical tool, \texttt{fast\_align} \cite{dummy}.
Doing so requires manually labelling the correspondences as either correct or incorrect.

@@

On the other hand, it is easy for shorter sentences to be assigned the minimum BLEU-1 to 4 score even when they only contain a single grammatical or semantic error. 
This is the case, for instance, of the sentence ``\textit{the test is oral}'', whose last word, ``\textit{oral}'' is translated as ``\textit{dura}'' (``\textit{hard}'') instead of ``\textit{orale}'' due to an alignment error. 
Nonetheless, it is worth noticing that this is one of the many cases in which the correct alignment $\langle$``\textit{oral}'',``\textit{orale}''$\rangle$ is also identified by the CE module, but hidden by the fact that, for simplicity, we only take into account the first translation candidate.
This suggest that, in contexts where higher precision is necessary, there are simple improvements that can be done either in the choice of the alignment criteria or at a later stage, in terms alignment selection. 
In other contexts, it can be more practical to aim for higher coverage and not perform any automatic selection, leaving lexicon refinements to a human, possibly with the help of the postprocessing tools that are currently under development \cite{dummy}.

>>>>

\section{Introduction}
Grammar-based translation pipelines such as those based on Grammatical Framework (GF) have been successfully employed in domain-specific Machine Translation (MT) \cite{ranta-etal-2020-abstract}.  
What makes these systems well suited to the task is the fact that, when we constrain ourselves to a specific domain, where precision is most often more important than coverage, they can provide strong guarantees in terms of grammatical correctness.

@@

This work is a new attempt in the same direction, enabled by two multilinguality-oriented grammar formalisms developed over the course of the last 25 years: Universal Dependencies (UD) \cite{ws-2019-universal} and Grammatical Framework (GF) \cite{ranta-etal-2020-abstract}.

@@

UD parsing alone would then be sufficient to extract (or propagate) tree-to-tree alignments, but not to automate the generation of a morphologically-aware translation lexicon for a generative grammar. 
This is where GF comes into play: after correspondences are inferred from a parallel text, our proposed system is able to convert them to GF grammar rules, easy to embed in a domain-specific grammar but also making it possible to immediately evaluate the system by carrying out small-scale translation experiments using pre-existing grammatical constructions implemented in GF's Resource Grammar Library (RGL). \todo{say what the RGL is} 
This is made possible by \texttt{gf-ud}, a conversion tool described in \cite{kolachina-ranta-2016-abstract} and \cite{ranta-kolachina-2017-universal}.

@@

The pseudocode of the basic extraction algorithm is shown in Figure \ref{TODO:}\todo{add pseudocode}.
The input is a list of priority-sorted \textit{alignment criteria}, i.e.rules to determine whether two dependency trees should be aligned with each other, which will be discussed in greater detail in Section \ref{criteria}, and UD tree representation of the pair of sentences to align.
From an implementation point of view, UD trees are, as in \texttt{gf-ud}, rose trees where each node represents a token, obtained from the CoNLL-U files produced by the UD parser.

@@

\paragraph{Part-Of-Speech equivalence}
CoNNL-U files provide information not only on the syntactic role of the tokens a sentence is composed of, but also on their grammatical categories, represented as Universal Part-Of-Speech (POS) tags \cite{petrov-etal-2012-universal}. 
Intuitively, if the nodes of two trees in matching contexts have the same POS tags, the two trees are more likely to correspond to each other than if not.

This is especially true if we focus, for instance, solely on the open class words (in the current implementation, defined as in the UD documentation \footnote{\url{universaldependencies.org/u/pos/all.html}}), thus ignoring prepositions, determiners, auxiliary verbs and all other function words, which tend to behave differently across different languages.

@@

\paragraph{Known translation divergence}
Parallel texts often present significant, systematic cross-linguistic grammatical distinctions. 
When this is the case, it is often straightforward to define alignment criteria based on recognizing the corresponding patterns.
If many distinctions of this kind are specific to particular language pairs or even stylistical, some of them occur independently of what languages are involved and have nothing to do with idiomatic usage or aspectual, discourse, domain or word knowledge. Drawing inspiration from \cite{dorr-1994-machine}, we refer to them as \textit{translation divergences} and handle the most common ones via alignment criteria that are used by default. \todo{add a few examples (<= one per type of divergence)}

@@

\section{Evaluation} \label{evaluation}
% TODO: move around things to have something here (or outline structure of the section)
\subsection{Data}
Because we want part of our evaluation to be independent from the quality of UD parsing, some of the experiments are carried out using manually annotated data. 
To this purpose, we use a 100-sentences subset of the Parallel UD (PUD) corpus, a set of 1000 handcrafted\footnote{Annotation is sometimes semi-automatic, but all treebanks are at least manually checked.}\todo{not even, I fear. Check this} multilingual treebanks in CoNLL-U format created for the CoNLL 2018 shared task on Multilingual Parsing from Raw Text to Universal Dependencies \cite{zeman-etal-2018-conll}.
PUD treebanks are available in over 20 languages, of which we selected English, Italian and Swedish. 
Using this data limits the amount of alignment errors that are due annotation issues to a minimum, even though a small number of inconsistencies and imprecisions is present even in this corpus.

When it comes to testing the program on raw text, we use two bilingual sentence-aligned corpora consisting of course plans from the Department of Mathematics and Computer Science of the University of Perugia (for English-Italian) and from the Department of Computer Science and Engineering (CSE) shared between the University of Gothenburg and the Chalmers University of Technology (for English-Swedish). 
For brevity, we will refer to these two datasets as to the DMI and the CSE corpora, or collectively as to the course plans datasets.
This data, available in the project repository, was collected and sentence-aligned specifically for this work and a related Bachelor's thesis project \cite{thesis}.
When using raw text, our parser of choice is UDPipe \cite{straka-etal-2016-udpipe}. 
In particular, we use the ParTUT English and Italian models for the DMI corpus and models trained on the bilingual LinES English-Swedish treebank for the CSE corpus.

@@

We assess the correctness of the alignments the CE module is able to extract first from manually annotated data, then from the DMI and CSE corpus, comparing our results with those obtained with a standard statistical tool, \texttt{fast\_align} \cite{dyer-etal-2013-simple}.
Doing so requires manually labelling the correspondences as either correct or incorrect.

@@

On the other hand, it is easy for shorter sentences to be assigned the minimum BLEU-1 to 4 score even when they only contain a single grammatical or semantic error. 
This is the case, for instance, of the sentence ``\textit{the test is oral}'', whose last word, ``\textit{oral}'' is translated as ``\textit{dura}'' (``\textit{hard}'') instead of ``\textit{orale}'' due to an alignment error. 
Nonetheless, it is worth noticing that this is one of the many cases in which the correct alignment $\langle$``\textit{oral}'',``\textit{orale}''$\rangle$ is also identified by the CE module, but hidden by the fact that, for simplicity, we only take into account the first translation candidate.
This suggest that, in contexts where higher precision is necessary, there are simple improvements that can be done either in the choice of the alignment criteria or at a later stage, in terms alignment selection. 
In other contexts, it can be more practical to aim for higher coverage and not perform any automatic selection, leaving lexicon refinements to a human, possibly with the help of the postprocessing tools that are currently under development.
